# Distributions

```{r packages}
library("ggplot2"); theme_set(theme_bw())
```

## Discrete

### Bernoulli

### Binomial

### Poisson

### Negative binomial

## Continuous

### Normal

### Logistic 

We say a random variable $X$ has a logistic distribution if it has the 
following probability density function. 
$$f(x) = \frac{e^{-(x-\mu)/\sigma}}{\sigma\left(1 + e^{-(x-\mu)/\sigma} \right)^2}$$
with 

- location parameter $\mu$,
- scale parameter $\sigma > 0$, and
- $-\infty < x < \infty$.

We write $X \sim Lo(\mu,\sigma)$ and $E[X] = \mu$ and $Var[X] = \sigma^2 \pi^2 / 3$. 

As can be seen in the plot below, 
the logistic distribution has much heavier tails than the normal distribution.

```{r logistic-pdf}
# Compare logistic and normal pdfs (same mean/variance)
ggplot(data.frame(x = c(-3, 3)),
       aes(x = x)) +
  stat_function(fun = dlogis, args = list(scale = 3 / pi^2)) + 
  stat_function(fun = dnorm, color = "red", linetype = "dotted") 
```

The logistic distribution gets is name from its cumulative distribution
function which is a member of the logistic functions.

$$F(x) = \frac{1}{1 + e^{-(x-\mu)/\sigma}}$$

The logistic distribution can be utilized in a latent variable representation
of logistic regression. 
